A Pandemic-Scale Ancestral Recombination Graph for SARS-CoV-2

### 2025-10-16 latex/main.tex
Comment in the discussion

Sc2ts reconstructs an ARG by incrementally adding daily batches of 
sequences (Figure~\ref{fig:methodology}).
First, sc2ts infers likely ``copying paths'' connecting each sample in the 
batch to node(s) in the current ARG
using a simplified version of the Li and Stephens (LS) model\cite{LiStephens2003}. 
% Important to go into a bit of detail about LS here as it's so fundamental
% and novel in the area.
The LS model is a Hidden Markov Model (HMM) widely used in human genomics
to approximate the effects of mutation and recombination on genetic
inheritance\citep{Delaneau2019-wl,Browning2018-nk},
typically parameterised by population-scaled mutation and
recombination rates varying along the genome~\cite{Donnelly2010-coalescent}.
In the interest of interpretability, we simplified this model to 
use a single parameter $k$ which essentially 
controls the number of recurrent mutations that is preferred to 
switching to a different parent via recombination (we use $k=4$ here
[TODO cite support for this, like CovRecomb etc]).
We used the highly efficient LS HMM implementation in tsinfer\cite{Kelleher2019},
with some additional optimisations enabled by our simplified model,
to find exact Viterbi solutions of these copying paths for each sample
given the current ARG.
Note that this is a form of parsimony, with the recombination
penalty $k$ defining the relative parsimoniousness of recombinant
vs non-recombinant solutions to the HMM.
The per-sample copying paths produced by the HMM (most of which do not 
involve recombination) form clusters of samples within a batch, which 
are then resolved using standard phylogenetic techniques.
Finally, sc2ts attaches the phylogenies of the clusters to the current ARG, and
applies parsimony-improving heuristics to address issues introduced 
by the inherent greediness of this strategy.
Sc2ts is implemented in Python,
building on core PyData infrastructure~\cite{Harris2020array,Virtanen2020scipy,
Mckinney2010data,Lam2015numba} and bioinformatics 
packages\cite{Shirley2015efficient,Kunzmann2018biotite,Kunzmann2023biotite}.
See STAR Methods for details on all aspects of the method.

@@

This estimate for effective recombination in SARS-CoV-2 is much lower 
than previous studies have suggested.
% Bolotie: , we identified 225 anomalous SARS-CoV-2 genomes of likely
% recombinant origins out of the first 87,695 genomes to be released
In an early analysis, 225 potentially recombinant sequences out 
of 87,695 samples [WHAT DATE?]\cite{Varabyou2021rapid}.
Over the same time period, XX events are present in the sc2ts ARG,
representing a total of XX samples.
In the flagship analysis for the RIPPLES method~\cite{Turakhia2022}, 
589 unique recombination events in a tree of 1.6 million samples
up to May 2021 were identified after extensive quality control.
These nodes had a combined total of 43,104 descendants, suggesting
that approximately 2.7\% of genomes were of recombinant origin.
Over the same time period, the sc2ts ARG contains 14 recombinants
passing QC with a total of 172 descendants. The ARG contains a total
501,235 samples over this time period, giving an overall total fraction of
0.034\%. Thus, the sc2ts ARG estimates 42-fold fewer recombination
events and 79-fold descendant samples over the same time period.
The numbers are not directly comparable as the published RIPPLES 
estimate is based on GISAID data which is enriched for
artefacts~\cite{Hunt2024}, but given the observations in 
the previous section on running RIPPLES on the Viridian data
it seems likely that RIPPLES is overestimating the prevalence
of recombination in SARS-CoV-2.

@@

We developed a SnpIt-like~\cite{OToole2024} visualisation 
that shows the informative genomic positions between 
a recombinant and its parent sequences, and highlights 
potentially problematic loci
(See Figure~\ref{fig:copying_pattern_examples} caption for details).
Inspection of these copying patterns for all the recombination events
identified associations between recombination events that failed QC and
known problematic loci in the SARS-CoV-2 genome.
Figure~\ref{fig:copying_pattern_examples} shows illustrative copying patterns
(Q1 to Q4 correspond to the labeled quadrants in Figure~\ref{fig:recombinant_qc}).
Q1 shows a well supported recombination event that passed QC (albeit marginally),
having a net number of supporting loci of 4 on both sides of the breakpoint.
Q2 shows an artefactual recombination event
caused by a 6-base Delta lineage-defining deletion.
There are 71 (14\%) QC-failing recombination events involving this deletion, and
many cases are affected by a known amplicon dropout
caused by this particular deletion \cite{Sanderson2021}.
Q3 shows an artefactual recombination event caused by
a combination of this Delta deletion and a known MNS (see below).
Finally, Q4 shows an artifactual recombination event
associated with the A507T-T508C-G509A MNS,
which has been identified as a sequencing error
due to incomplete read trimming \cite{DeMaio2024}.
There are 9 recombination events involving this particular MNS,
of which 6 failed QC.

>>>>

Sc2ts reconstructs an ARG by incrementally adding daily batches of 
sequences (Figure~\ref{fig:methodology}).
First, sc2ts infers likely ``copying paths'' connecting each sample in the 
batch to node(s) in the current ARG
using a simplified version of the Li and Stephens (LS) model\cite{LiStephens2003}. 
% Important to go into a bit of detail about LS here as it's so fundamental
% and novel in the area.
The LS model is a Hidden Markov Model (HMM) widely used in human genomics
to approximate the effects of mutation and recombination on genetic
inheritance\citep{Delaneau2019-wl,Browning2018-nk},
typically parameterised by population-scaled mutation and
recombination rates varying along the genome~\cite{Donnelly2010-coalescent}.
In the interest of interpretability, we simplified this model to 
use a single parameter $k$ which essentially 
controls the number of recurrent mutations that is preferred to 
switching to a different parent via recombination 
% This is quite a big thing really, and ultimately what determines
% the number of recombinants detected.
(we use $k=4$ here; see STAR Methods and Discussion).
We used the highly efficient LS HMM implementation in tsinfer\cite{Kelleher2019},
with some additional optimisations enabled by our simplified model,
to find exact Viterbi solutions of these copying paths for each sample
given the current ARG.
Note that this is a form of parsimony, with the recombination
penalty $k$ defining the relative parsimoniousness of recombinant
vs non-recombinant solutions to the HMM.
The per-sample copying paths produced by the HMM (most of which do not 
involve recombination) form clusters of samples within a batch, which 
are then resolved using standard phylogenetic techniques.
Finally, sc2ts attaches the phylogenies of the clusters to the current ARG, and
applies parsimony-improving heuristics to address issues introduced 
by the inherent greediness of this strategy.
Sc2ts is implemented in Python,
building on core PyData infrastructure~\cite{Harris2020array,Virtanen2020scipy,
Mckinney2010data,Lam2015numba} and bioinformatics 
packages\cite{Shirley2015efficient,Kunzmann2018biotite,Kunzmann2023biotite}.
See STAR Methods for details on all aspects of the method.

@@

The number of recombination events that we detect is largely 
driven by the choice of recombination penalty, $k$ (or $p$
for RIPPLES). 
At $k=4$ we detect fewer recombinants overall than the combination of 
UShER and RIPPLES at the same threshold. 
However, most of the extra recombinants detected by UShER+RIPPLES
involve violations of time ordering; whether this is due to incorrect
metadata or these are false positives would require further study.
However, at $p=4$, UShER+RIPPLES detects very few of the well-characterised
Pango X-lineages which are detected by sc2ts. At the more lenient
threshold of $p=3$, more of these events are detected (but still fewer
than sc2ts) but at the cost of nearly 4 times more recombination
events overall. Our detailed analysis shows that RIPPLES has a well
defined task, which it performs well, finding the potential events
in an exhaustive search. However, it can only operate within the
input that it is provided, and it is the UShER tree provided with
the Viridian data that does not have clear signals for recombination
of X lineages such as XBB at the $p=4$ threshold. 
The inference of recombination is therefore intrinsically linked
with the underlying genealogy inference method.

[Whether k=3 or k=4 is ``correct'' is a matter for debate, and feeds
into the next point.]

The estimate for effective recombination in SARS-CoV-2 that we obtain
at k=4 is much lower 
than previous studies have suggbalanced
% Bolotie: , we identified 225 anomalous SARS-CoV-2 genomes of likely
% recombinant origins out of the first 87,695 genomes to be released
In an early analysis, 225 potentially recombinant sequences out 
of 87,695 samples collected before October 2020\cite{Varabyou2021rapid}.
Over the same time period, XX events are present in the sc2ts ARG,
representing a total of XX samples.
In the flagship analysis for the RIPPLES method~\cite{Turakhia2022}, 
589 unique recombination events in a tree of 1.6 million samples
up to May 2021 were identified after extensive quality control.
These nodes had a combined total of 43,104 descendants, suggesting
that approximately 2.7\% of genomes were of recombinant origin.
Over the same time period, the sc2ts ARG contains 14 recombinants
passing QC with a total of 172 descendants. The ARG contains a total
501,235 samples over this time period, giving an overall total fraction of
0.034\%. Thus, the sc2ts ARG estimates 42-fold fewer recombination
events and 79-fold descendant samples over the same time period.
The numbers are not directly comparable as the published RIPPLES 
estimate is based on GISAID data which is enriched for
artefacts~\cite{Hunt2024}, but given the observations in 
the previous section on running RIPPLES on the Viridian data
it seems likely that RIPPLES is overestimating the prevalence
of recombination in SARS-CoV-2.

@@

We developed a SnipIt-like~\cite{OToole2024} visualisation 
that shows the informative genomic positions between 
a recombinant and its parent sequences, and highlights 
potentially problematic loci
(See Figure~\ref{fig:copying_pattern_examples} caption for details).
Inspection of these copying patterns for all the recombination events
identified associations between recombination events that failed QC and
known problematic loci in the SARS-CoV-2 genome.
Figure~\ref{fig:copying_pattern_examples} shows illustrative copying patterns
(Q1 to Q4 correspond to the labeled quadrants in Figure~\ref{fig:recombinant_qc}).
Q1 shows a well supported recombination event that passed QC (albeit marginally),
having a net number of supporting loci of 4 on both sides of the breakpoint.
Q2 shows an artefactual recombination event
caused by a 6-base Delta lineage-defining deletion.
There are 71 (14\%) QC-failing recombination events involving this deletion, and
many cases are affected by a known amplicon dropout
caused by this particular deletion \cite{Sanderson2021}.
Q3 shows an artefactual recombination event caused by
a combination of this Delta deletion and a known MNS (see below).
Finally, Q4 shows an artifactual recombination event
associated with the A507T-T508C-G509A MNS,
which has been identified as a sequencing error
due to incomplete read trimming \cite{DeMaio2024}.
There are 9 recombination events involving this particular MNS,
of which 6 failed QC.
